---
title: "Analysis of GCNT2 Expression Profile"
author: A. Mihaylov <aleksandar.mihaylov@epfl.ch>
---

# Preparing Finely Annotated Dataset

## Downloading Finely Annotated Dataset

```{r downloading_processed_dataset}
library(dplyr)

ds_name <- "SalasA_2023_GSE214695"
home.path <- file.path(getwd(), ds_name)
output_data_path <- file.path(home.path, "output")

size <- "whole" # "light" or "whole"

# cmc stands for colonic mucosa cells
cmc <- readRDS(file.path(output_data_path, paste0(ds_name, 
                                                  sprintf("_finely_annotated_%s.rds", size))))
```

## Extracting GCNT2 Counts 

```{r extracting_counts}
counts <- Seurat::GetAssayData(cmc, assay = "RNA", slot = "data")
gcnt2_counts <- counts["GCNT2", ]
cmc$GCNT2 <- gcnt2_counts
```

# Initial Analysis per Cell Type

## ANOVA Test

```{r anova_test}
cat("\033[31mAll Cells Considered\033[0m\n")

cat("\033[32mANOVA Table\033[0m\n")
gcnt2.cell_type.expression <- cmc@meta.data[, c("manual_annotation", "GCNT2")]
gcnt2.aov.model <- aov(gcnt2.cell_type.expression$GCNT2 ~ gcnt2.cell_type.expression$manual_annotation)
print(summary(gcnt2.aov.model))
cat("-----------------------------------------","\n")

cat("\033[32mTukeyHSD\033[0m\n")
gcnt2.tukey.results <- TukeyHSD(gcnt2.aov.model)
print(gcnt2.tukey.results)

cell_types <- unique(gcnt2.cell_type.expression$manual_annotation)
groups <- combn(cell_types, 2, simplify = FALSE)
gcnt2.wilcox.results <- purrr::map_dfr(groups, ~{
  group1 <- .x[1]
  group2 <- .x[2]
  
  data1 <- gcnt2.cell_type.expression %>%
    dplyr::filter(manual_annotation == group1) %>%
    dplyr::pull(GCNT2)
  
  data2 <- gcnt2.cell_type.expression %>%
    dplyr::filter(manual_annotation == group2) %>%
    dplyr::pull(GCNT2)
  
  test <- wilcox.test(data1, data2)
  
  tibble(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
})
print(gcnt2.wilcox.results)
```

## Initial Diagnostic Plots

```{r in_diganostic_plots}
Seurat::VlnPlot(cmc, features = "GCNT2", group.by = "manual_annotation") + 
  ggplot2::ggtitle("Violin Plots of GCNT2 Expression across Cell Types")

ggplot2::ggplot(gcnt2.cell_type.expression, 
                ggplot2::aes(x = manual_annotation, y = GCNT2)) + 
  ggplot2::geom_boxplot(fill="slateblue", alpha=0.2) + 
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("Boxplot of GCNT2 Expression across Cell Types")

Seurat::RidgePlot(cmc, features = "GCNT2", group.by = "manual_annotation") +
  ggplot2::ggtitle("Ridge Plot of GCNT2 Expression across Cell Types")

Seurat::FeaturePlot(cmc, features = "GCNT2") +
  ggplot2::ggtitle("Feature Plot of GCNT2 Expression across Cell Types")
```

Clearly, most cells, independent of their type, do not express GCNT2. Therefore, for downstream analyses, let us focus solely on the cells which do express GCNT2.

```{r subsetting_cells}
filtered_cmc <- subset(cmc, subset = GCNT2 > 0)
cat("Number of cells left in Seurat object:", length(Seurat::Cells(filtered_cmc)))
```

# Analysis per Cell Type 

## Statistical Tests

```{r stat_tests}
cat("\033[31mOnly GCNT2 Expressing Cells Considered\033[0m\n")

cat("\033[32mANOVA Table\033[0m\n")
gcnt2.cell_type.expression <- filtered_cmc@meta.data[, c("manual_annotation", "GCNT2")]
gcnt2.aov.model <- aov(gcnt2.cell_type.expression$GCNT2 ~ gcnt2.cell_type.expression$manual_annotation)
print(summary(gcnt2.aov.model))
cat("-----------------------------------------","\n")

cat("\033[32mTukey HSD\033[0m\n")
gcnt2.tukey.results <- TukeyHSD(gcnt2.aov.model)
print(gcnt2.tukey.results)
cat("-----------------------------------------","\n")

cat("\033[32mWilcoxon Test\033[0m\n")
gcnt2.wilcox.results <- wilcox.test(gcnt2.cell_type.expression)
```

## Diagnostic Plots

```{r plots_per_cell_type}
Seurat::VlnPlot(filtered_cmc, "GCNT2", group.by = "manual_annotation") +
  ggplot2::ggtitle("Violin Plots of GCNT2 Expression across Cell Types Omitting GCNT2 Non-expressing Cells") 

ggplot2::ggplot(gcnt2.cell_type.expression, 
                ggplot2::aes(x = manual_annotation, y = GCNT2)) + 
  ggplot2::geom_boxplot(fill="slateblue", alpha=0.2) + 
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("Boxplot of GCNT2 Expression across Cell Types Omitting GCNT2 Non-expressing Cells")

Seurat::RidgePlot(filtered_cmc, features = "GCNT2", group.by = "manual_annotation") +
  ggplot2::ggtitle("Ridge Plot of GCNT2 Expression across Cell Types Omitting GCNT2 Non-expressing Cells")
```

## Significance Plots

```{r significance_plots}
library(dplyr)

ggplot2::ggplot(gcnt2.cell_type.expression, ggplot2::aes(x = manual_annotation, y = GCNT2)) +
  ggplot2::geom_jitter(width = 0.05, size = 1) +
  ggplot2::stat_summary(fun = mean, geom = "crossbar", width = 0.5, 
                        fatten = 2, color = "magenta") +
  ggpubr::stat_compare_means(
    comparisons = list(combn(unique(gcnt2.cell_type.expression$manual_annotation), 2)),
    method = "wilcox.test",
    label = "p.signif") +
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("GCNT2 Expression vs. Cell Types Omitting GCNT2 Non-expressing Cells, Wilcoxon Test") +
  ggplot2::theme_classic()

gcnt2.tukey.results[[1]] %>% 
  as.data.frame() %>% 
  dplyr::rename(p_adj = `p adj`) %>% 
  dplyr::filter(p_adj <= 1 - attr(gcnt2.tukey.results, "conf.level")) %>% 
  tibble::rownames_to_column(var = "groups") %>% 
  tidyr::separate(groups, into = c("group1", "group2"), sep = "-") %>% 
  dplyr::mutate(y_pos = seq(from = 1.2 * max(gcnt2.cell_type.expression$GCNT2), 
                            to = 1.2 * max(gcnt2.cell_type.expression$GCNT2) + 0.7*nrow(.),
                            length.out = nrow(.))) %>%
  dplyr::mutate(p.sig = case_when(
    p_adj <= 0.0001 ~ "****",
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01 ~ "**",
    TRUE ~ "*",
  )) -> gcnt2.tukey.df

ggplot2::ggplot(gcnt2.cell_type.expression, ggplot2::aes(x = manual_annotation, y = GCNT2)) +
  ggplot2::geom_jitter(width = 0.05, size = 1) +
  ggplot2::stat_summary(fun = mean, geom = "crossbar", width = 0.5, 
                        fatten = 2, color = "magenta") +
  ggpubr::stat_pvalue_manual(
    data = gcnt2.tukey.df,
    label = "p.sig",
    y.position = "y_pos"
  ) +
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("GCNT2 Expression vs. Cell Types Omitting GCNT2 Non-expressing Cells, Tukey Test")
  ggplot2::theme_classic()
```

# Analysis per Condition

## Statistical Tests

```{r stat_tests}
cat("\033[31mAll Cells Considered\033[0m\n")

cat("\033[32mANOVA Table\033[0m\n")
gcnt2.condition.expression <- cmc@meta.data[, c("Treatment", "GCNT2")]
gcnt2.aov.model <- aov(gcnt2.condition.expression$GCNT2 ~ gcnt2.condition.expression$Treatment)
print(summary(gcnt2.aov.model))
cat("-----------------------------------------","\n")

cat("\033[32mTukey HSD\033[0m\n")
gcnt2.tukey.results <- TukeyHSD(gcnt2.aov.model)
print(gcnt2.tukey.results)
cat("-----------------------------------------","\n")

cat("\033[32mWilcoxon Test\033[0m\n")
gcnt2.wilcox.results <- wilcox.test(gcnt2.condition.expression)
cat("---------------------------------------------------------------------------------------------------","\n")

cat("\033[31mOnly GCNT2 Expressing Cells Considered\033[0m\n")

cat("\033[32mANOVA Table\033[0m\n")
gcnt2.condition.expression <- filtered_cmc@meta.data[, c("Treatment", "GCNT2")]
gcnt2.aov.model <- aov(gcnt2.condition.expression$GCNT2 ~ gcnt2.condition.expression$Treatment)
print(summary(gcnt2.aov.model))
cat("-----------------------------------------","\n")

cat("\033[32mTukey HSD\033[0m\n")
gcnt2.tukey.results <- TukeyHSD(gcnt2.aov.model)
print(gcnt2.tukey.results)
cat("-----------------------------------------","\n")

cat("\033[32mWilcoxon Test\033[0m\n")
gcnt2.wilcox.results <- wilcox.test(gcnt2.condition.expression)
```

## Diagnostic Plots

```{r plots_per_condition}
Seurat::VlnPlot(filtered_cmc, "GCNT2", group.by = "Treatment") +
  ggplot2::ggtitle("Violin Plots of GCNT2 Expression across Conditions Omitting GCNT2 Non-expressing Cells") 

ggplot2::ggplot(gcnt2.condition.expression, 
                ggplot2::aes(x = Treatment, y = GCNT2)) + 
  ggplot2::geom_boxplot(fill="slateblue", alpha=0.2) + 
  ggplot2::xlab("Condition") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("Boxplot of GCNT2 Expression across Conditions Omitting GCNT2 Non-expressing Cells")

Seurat::RidgePlot(filtered_cmc, features = "GCNT2", group.by = "Treatment") +
  ggplot2::ggtitle("Ridge Plot of GCNT2 Expression across Conditions Omitting GCNT2 Non-expressing Cells")
```

## Significance Plots

```{r significance_plots}
library(dplyr)

ggplot2::ggplot(gcnt2.cell_type.expression, ggplot2::aes(x = manual_annotation, y = GCNT2)) +
  ggplot2::geom_jitter(width = 0.05, size = 1) +
  ggplot2::stat_summary(fun = mean, geom = "crossbar", width = 0.5, 
                        fatten = 2, color = "magenta") +
  ggpubr::stat_compare_means(
    comparisons = list(combn(unique(gcnt2.cell_type.expression$manual_annotation), 2)),
    method = "wilcox.test",
    label = "p.signif") +
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("GCNT2 Expression vs. Cell Types Omitting GCNT2 Non-expressing Cells, Wilcoxon Test") +
  ggplot2::theme_classic()

gcnt2.tukey.results[[1]] %>% 
  as.data.frame() %>% 
  dplyr::rename(p_adj = `p adj`) %>% 
  dplyr::filter(p_adj <= 1 - attr(gcnt2.tukey.results, "conf.level")) %>% 
  tibble::rownames_to_column(var = "groups") %>% 
  tidyr::separate(groups, into = c("group1", "group2"), sep = "-") %>% 
  dplyr::mutate(y_pos = seq(from = 1.2 * max(gcnt2.cell_type.expression$GCNT2), 
                            to = 1.2 * max(gcnt2.cell_type.expression$GCNT2) + 0.7*nrow(.),
                            length.out = nrow(.))) %>%
  dplyr::mutate(p.sig = case_when(
    p_adj <= 0.0001 ~ "****",
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01 ~ "**",
    TRUE ~ "*",
  )) -> gcnt2.tukey.df

ggplot2::ggplot(gcnt2.cell_type.expression, ggplot2::aes(x = manual_annotation, y = GCNT2)) +
  ggplot2::geom_jitter(width = 0.05, size = 1) +
  ggplot2::stat_summary(fun = mean, geom = "crossbar", width = 0.5, 
                        fatten = 2, color = "magenta") +
  ggpubr::stat_pvalue_manual(
    data = gcnt2.tukey.df,
    label = "p.sig",
    y.position = "y_pos"
  ) +
  ggplot2::xlab("Cell Type") +
  ggplot2::ylab("GCNT2 Expression") +
  ggplot2::ggtitle("GCNT2 Expression vs. Cell Types Omitting GCNT2 Non-expressing Cells, Tukey Test")
  ggplot2::theme_classic()
```


